---
title: "ICEWS Aggregation"
author: "Richard Neureuther"
format: html
editor: visual
---

# Cleaning and aggregation pipeline of ICEWS event data

### set working directory

```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = "C:/Users/Richard Neureuther/OneDrive/Documents/SEDS/Master Thesis/Data"
)

```

### Library calls

```{r}
library(tidyverse)
library(countrycode)
```

### Load country lookup and define helper function

```{r}
# Load country lookup once
countries <- read.csv("VIEWS/countires_reduced.csv", sep = ";")

# Helper: rename ICEWS columns once
rename_icews_columns <- function(df) {
  colnames(df) <- c(
    "eventID", "event_date", "source_name", "source_sector", "source_country",
    "text", "CAMEO", "intensity", "target_name", "target_sectors",
    "target_country", "storyID", "Num_sentence", "puplisher", "city",
    "district", "provice", "country", "latitude", "longitude"
  )
  df
}
```

## Mapping for CAMEO codes to aggregated variables

I do the mapping explicitly so that I can double check that all cameo codes that should be included/excluded are correctly specified. (CAMEO 057 is removed from Var 1 and Var 8 to prevent join issues, will be handled separately)

```{r}
# CAMEO map once
cameo_map <- bind_rows(
  #Variable 1: De-escalation
  tibble(CAMEO = c("0355","0356","0357","046","050","054","055","056",
                   "082","087","0871","0872","0873","0874"),
         variable = "deescalation"),
  
  #Variable 2: Breakdown 
  tibble(CAMEO = c("120","121","1211","1212","1213","122","123","1231","1232","1233","1234","124","125","126","128",
                   "160","161","162","1621","1622","1623","163","164","1641","1642","1643","165","166"),
         variable = "breakdown"),
  
  #Variable 3: Protest 
  tibble(CAMEO = c("140","141","1411","1412","1413","1414","142","1421","1422","1423","1424",
                   "143","1431","1432","1433","1434","144","1441","1442","1443","1444",
                   "145","1451","1452","1453","1454"),
         variable = "protest"),
  
  #Variable 4: Repression intensification 
  tibble(CAMEO = c("170","171","1711","1712","172","1721","1722","1723","1724","173","174","175"),
         variable = "repression_up"),
  
  #Variable 5: Repression easing 
  tibble(CAMEO = c("034","0341","0342","0343","0344",
                   "035","0351","0352","0353","0354",
                   "036","037","038","039",
                   "080","081","0811","0812","0813","0814",
                   "083","0831","0832","0833","0834",
                   "084","0841","0842","085"),
         variable = "repression_down"),
  
  #Variable 6: Pre-violence escalation
  tibble(CAMEO = c("104","1041","1042","1043","1044","106",
                   "130","131","1311","1312","1313",
                   "132","1321","1322","1323","1324",
                   "133","134","135","136","137",
                   "138","1381","1382","1383","1384","1385",
                   "139","150","151","152","153","154"),
         variable = "escalation_nonviolent"),
  
  #Variable 7: Non-lethal escalation
  tibble(CAMEO = c("180","181","185","190","191","192","193","196"),
         variable = "escalation_nonlethal"),
  
  #Variable 8: External involvement
  tibble(CAMEO = c("045","060","061","062","063","064",
                   "070","071","072","073","074",
                   "086","0861","0862","0863"),
         variable = "external")
)

```

## Define Pipeline

This part creates one big functions that handles the entire preprocessing and aggregation for the ICEWS data The function is created from multiple smaller function defined in an earlier iteration of the project. I decided to handle everything in one large function, so that I can iterate over each year of the ICEWS events and aggregate the entire dataset with a few clicks.

```{r}
#preprocessing function 
process_icews_year <- function(year, data_dir, countries, cameo_map) {
  # create file path for the year
  fpath <- file.path(data_dir, paste0("events_", year, ".tab"))
  
  #load  raw ICEWS data frame 
  df <- read_tsv(fpath, show_col_types = FALSE) %>%
    #apply renaming function defined earlier 
    rename_icews_columns() %>%
    #standardize CAMEO column 
    mutate(

      CAMEO = as.character(CAMEO),
      CAMEO = str_trim(CAMEO),
      #after 2020, the ICEWS team changed their classification convention and removed the
      #leading zero from single digit CAMEO codes (1-9)
      #I check whether the CAMEO code has length 2 (i.e., is missing leading zero)
      #and add it back if necessary
      CAMEO = if_else(
        nchar(CAMEO) == 2,
        str_pad(CAMEO, width = 3, side = "left", pad = "0"),
        CAMEO
      ),
      
      #trim white space from country columns
      source_country = str_trim(source_country),
      target_country = str_trim(target_country),
      country        = str_trim(country)
    ) %>%
    
    #remove rows where country is NA (classification mistakes)
    filter(!is.na(country)) %>%
    
    #add iso3 label to each row 
    mutate(
      iso3c = countrycode(
        sourcevar   = country,
        origin      = "country.name",
        destination = "iso3c",
        warn        = FALSE
      )
    ) %>%
    
    #keep only countries that are in the country lookup table 
    filter(iso3c %in% countries$isoab) %>%
    
    #remove all non domestic events exeptfor events relevant for international involvement 
    filter(
      source_country == target_country |
        (source_country != target_country &
           (grepl("^(06|07)", CAMEO) |
              CAMEO %in% c("045","057","086","0861","0862","0863")))
    ) %>%
  
    #remove irrelevant cameo codes
    filter(!grepl("^(01|02|09|11|20)", CAMEO)) %>%
    filter(!grepl("^(1823|183|184|186)", CAMEO)) %>%
    
    #transform event_date to Date type and extract year and month
    mutate(
      event_date = as.Date(event_date),
      year  = year(event_date),
      month = month(event_date)
    ) %>%
    
    #add variable labels to each event row using the CAMEO map
    left_join(cameo_map, by = "CAMEO") %>%
    
    #hadle CAMEO 057 separately
    #assign  Var 1 if domestic, Var 8 if international 
    mutate(
      variable = case_when(
        CAMEO == "057" & source_country != target_country ~ "external",
        CAMEO == "057" & source_country == target_country ~ "deescalation",
        TRUE ~ variable
      )
    ) %>%
    
    #remove all row that have irrelevant event types (i.e., variable is NA)
    filter(!is.na(variable))
  
  #Robustness check. Show number of event rows per month (used later)
  month_event_counts <- df %>%
  count(year, month, name = "n_event_rows")

  # change to wide format with counts per country-month-variable
  wide <- df %>%
    group_by(country, iso3c, year, month, variable) %>%
    summarise(n_events = n(), .groups = "drop") %>%
    pivot_wider(
      names_from  = variable,
      values_from = n_events,
      values_fill = list(n_events = 0)
    )

  #Ensure all expected columns exist (important if a year has zero events in a category)
  expected_vars <- c(
    "deescalation","breakdown","protest","repression_up",
    "repression_down","escalation_nonviolent","escalation_nonlethal","external"
  )
  
  #find missing columns
  missing_vars <- setdiff(expected_vars, names(wide))
  #add missing columns and fill with zeros
  if (length(missing_vars) > 0) {
    wide[missing_vars] <- 0L
  }

  #construct an (empty) complete country-month panel 
  panel_skeleton <- wide %>%
    distinct(country, iso3c, year) %>%
    expand_grid(month = 1:12)
  
  # Merge observed data onto full panel and replace NAs with zeros
  panel <- panel_skeleton %>%
    left_join(wide, by = c("country", "iso3c", "year", "month")) %>%
    mutate(across(all_of(expected_vars), ~ replace_na(.x, 0L))) %>%
    arrange(iso3c, year, month)
  
  #return completed country month panel and month event counts for robustness check
  list(
  panel = panel,
  month_event_counts = month_event_counts
)
}
```

### Handle weekly data for 2023

For the pipeline to work smoothly I now import the weekly data for 2023, concatenate it and save it into the directory as a tab file

```{r}
#set directory for 2023 weekly data
dir <- "ICEWS/dataverse_files_unzipped/2023"

#fetch all file names in the folder and store them
files <- list.files(
  path = dir,
  pattern = "icews-events\\.tab$",
  full.names = TRUE
) |> sort()

#read and concatenate all weekly files into one data frame
events_2023 <- files %>%
  map_dfr(
    read_tsv,
    col_types = cols(`CAMEO Code` = col_character()),
    show_col_types = FALSE
  )

#remove all rows form 2022 (for some reason the ICEWS team included data from 2022 in the 2023 weekly data)
events_2023 <- events_2023 %>%
  filter(!str_starts(`Event Date`, "2022"))
```

```{r}
#save concatenated 2023 data as tab file for later processing
write_tsv(
  events_2023,
  file = file.path("ICEWS/dataverse_files_unzipped/events_2023.tab")
)
```

### Apply the pipeline to all years and combine into one data frame

Here I apply the preprocessing and aggregation pipeline to each year of the event data separately. The yearly panels are then merge into one data frame containing all years from 1995 to 2022.

```{r}
#set directory and year range 
data_dir <- "ICEWS/dataverse_files_unzipped"
years <- 1995:2023

#create list for each panel and initialize names
results <- vector("list", length(years))
names(results) <- as.character(years)

#iterate over each year 
for (i in seq_along(years)) {
  y <- years[i]
  #print progress message 
  message("Processing year: ", y)
  #apply preprocessing and aggregation function
  results[[i]] <- process_icews_year(
    year      = y,
    data_dir  = data_dir,
    countries = countries,
    cameo_map = cameo_map
  )
}

#combine all yearly panels into one data frame
icews_full_panel <- bind_rows(lapply(results, `[[`, "panel"))

# enforce integer type for count columns
count_cols <- c(
  "deescalation","breakdown","protest","repression_up",
  "repression_down","escalation_nonviolent","escalation_nonlethal","external"
)

icews_full_panel <- icews_full_panel %>%
  mutate(across(all_of(count_cols), as.integer))


month_event_counts_all <- bind_rows(lapply(results, `[[`, "month_event_counts"))
```

-\> warnings for 2017 are due to some NULL values in latitude/longitude columns that are not relevant for my analysis

## Robustness checks for the aggregated data frame

Here I now perform some basic checks to ensure that the aggregation was successful and we have no obvious issues with the data.

### Check if there are any completely empty months (i.e. no relevant events occured)

```{r}
empty_months <- expand_grid(year = years, month = 1:12) %>%
  left_join(month_event_counts_all, by = c("year", "month")) %>%
  filter(is.na(n_event_rows)) %>%
  arrange(year, month)

if (nrow(empty_months) == 0) {
  message("There are no completely empty months.")
} else {
  empty_months
}
```

-\> as expected 2023-05 onwards is missing

### Remove april 2023 onwards

Since we only have partly data for April of 2023 and no data after that, I remove those months completely. April is removed because the model will handle data monthly, leaving april in would confuse learning as the n of events in april are underrepresented.

```{r}
icews_full_panel <- icews_full_panel %>%
  filter(!(year == 2023 & month >= 4))
```

### Check if there are any duplicates

```{r}
duplicates <- icews_full_panel %>%
  count(iso3c, country, year, month) %>%
  filter(n > 1)

#print duplicates if there are any
if (nrow(duplicates) == 0) {
  message("No duplicates found.")
} else {
  duplicates
}

```

### check if each country year has 12 entries

```{r}
short_months <- icews_full_panel %>%
  filter(year != 2023) %>% #exclude 2023 as we only have data till march
  count(iso3c, year) %>%
  filter(n != 12)

if (nrow(short_months) == 0) {
  message("All country-years have 12 months.")
} else {
  short_months
}

```

### Check for NAs in count variables

```{r}
expected_vars <- c(
  "deescalation","breakdown","protest","repression_up",
  "repression_down","escalation_nonviolent","escalation_nonlethal","external"
)

na_counts <- icews_full_panel %>%
  summarise(across(all_of(expected_vars), ~ sum(is.na(.x)))) 

na_counts

```

-\> no NAs found

### check if all years are present and if there are any years outside the 1995 to 2023 range

```{r}
years_present <- icews_full_panel %>%
  distinct(year) %>%
  arrange(year)

years_present

bad_years  <- icews_full_panel %>% filter(year < 1995 | year > 2023)

bad_years

```

### summarise all observations

```{r}
panel_year_totals <- icews_full_panel %>%
  group_by(year) %>%
  summarise(across(all_of(expected_vars), sum, na.rm = TRUE), .groups = "drop")
panel_year_totals

```

#export dataset as csv

```{r}
write_csv(
  icews_full_panel,  
  file = "ICEWS/aggregated_files/icews_aggregated.csv"
)
```

## EDA

Create a long format of the dataset for visualisation
```{r}
expected_vars <- c(
  "deescalation","breakdown","protest","repression_up",
  "repression_down","escalation_nonviolent","escalation_nonlethal","external"
)

panel_year_long <- icews_full_panel %>%
  group_by(year) %>%
  summarise(across(all_of(expected_vars), \(x) sum(x, na.rm = TRUE)), .groups = "drop") %>%
  pivot_longer(
    cols = all_of(expected_vars),
    names_to = "variable",
    values_to = "n_events"
  )
```

```{r}

pretty_labels <- c(
  deescalation            = "Deescalation",
  breakdown               = "Breakdown & Intransigence",
  protest                 = "Mobilization & Dissent",
  repression_up           = "Increasing Repression",
  repression_down         = "Decreasing Repression",
  escalation_nonviolent   = "Pre Violence Escalation",
  escalation_nonlethal    = "Nonlethal Escalation",
  external                = "External Involvement"
)
agg_plot <- ggplot(panel_year_long,
                   aes(x = year, y = n_events, color = variable)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_discrete(
    name   = "Event type",
    labels = pretty_labels
  ) +
  labs(
    x = "Year",
    y = "Number of events",
    title = "Frequency of Aggregated ICEWS Events Over Time"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(hjust = 0.7, face = "bold"),
    legend.title = element_text(face = "bold")
  )

```

```{r}
agg_plot
```

#separate plots for each variable

```{r}
sep_plot <- ggplot(panel_year_long, aes(x = year, y = n_events)) +
  geom_line(linewidth = 1, color = "steelblue") +
  facet_wrap(
    ~ variable,
    scales = "free_y",
    labeller = labeller(variable = pretty_labels)
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Year",
    y = "Number of events",
    title = "Frequency of Aggregated ICEWS Events Over Time"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.45, face = "bold"),
    strip.text = element_text(face = "bold")
  )


```

```{r}
sep_plot
```
Explanation for sharp increase in mobilization and Dissen variable:
Mobilization and dissent events increase sharply in 2010–2011, coinciding with the onset of the Arab Spring, which has been widely identified as a catalyst for mass protest and regional mobilization (Beissinger et al., 2015). (https://www.jstor.org/stable/43664167?seq=1)

export plots with high dpi

```{r}
ggsave(
  filename = "ICEWS/aggregated_files/icews_event_categories_over_time_agg.png",
  plot     = agg_plot,
  width    = 10,
  height   = 6,
  dpi      = 300
)

ggsave(
  filename = "ICEWS/aggregated_files/icews_event_categories_over_time_sep.png",
  plot     = sep_plot,
  width    = 10,
  height   = 6,
  dpi      = 300
)
```
